---
title: "Fitabase"
author: "Olena Urbanska"
date: "2023-12-07"
output: html_document
---


## R Markdown

#

Let's start with loading libraries for work 
```{r}
library(tidyverse)
library(dplyr)
library(readr)
library(lubridate)
library(ggplot2)
library(tidyr)
library(janitor)
```


Firstly, I've decided to explore information about daily rates, because with them we can figure out how many people from our sample use this device day-to-day. 

Let's upload first data set "daily_steps" and see it's structure. Then we'll perform cleaning process.


```{r}
daily_steps1 <- read_csv("dailySteps_merged.csv")
daily_steps <- clean_names(daily_steps1)
head(daily_steps)

```

This data set contains 3 variable "id", "activity_day", "step_total". All 33 users 
have used this build in feature.
```{r}
n_distinct(daily_steps$id)
                          
                      
```

Let's upload and examine data set "dailyActivity_merged.csv" to look through users 
daily activity.

```{r}
daily_activity1 <- read_csv("dailyActivity_merged.csv")
daily_activity <- clean_names(daily_activity1)
glimpse(daily_activity)
```

In this step we find out average value for steps, distance and loss of calories.
This will help us to understand is there a correlation between average steps fixed by 
smart device and loss of calories. 


```{r}
steps_calories <- daily_activity %>% 
  select(id, total_steps, total_distance, calories) %>% group_by(id) %>% 
  summarize(avg_steps = mean(total_steps),
             avg_distance = mean(total_distance),
            avg_calories = mean(calories)) %>% arrange(avg_steps,avg_calories)
head(steps_calories)
```


```{r}
ggplot(data = steps_calories, mapping = aes(x = avg_steps, y = avg_calories)) + 
  geom_point() +
  geom_smooth()
```
With this visualization we can see that it is positive correlation between average steps and loosing of calories. So , we have found this trend , that people who walk more tend to lose more calories.

As there is sometimes missing values, we replace 0 values as NA in column total_steps to make our calculation.
```{r}
daily_activity$total_steps[daily_activity$total_steps == 0] <- NA
```

Let's calculate information about average steps of all our users.

```{r}
steps_info <- daily_activity %>% 
  summarize(min_steps = min(total_steps, na.rm = TRUE),
            avg_steps = mean(total_steps, na.rm = TRUE),
            medium_steps = median(total_steps, na.rm = TRUE),
            max_steps = max(total_steps, na.rm = TRUE),
            standart_deviation = sd(total_steps, na.rm = TRUE))
steps_info
```
The mean (average) of a data set is found by adding all numbers in the data set and then dividing by the number of values in the set. The median is the middle value when a data set is ordered from least to greatest [link](https://www.khanacademy.org/)

Most pedometers provide guidelines as per the 10,000-step protocol. You can, however, tweak your routine to clock 7,500 steps daily. Pedometers classify activity as follows:

Sedentary: Less than 5,000 steps daily
Low active: About 5,000 to 7,499 steps daily
Somewhat active: About 7,500 to 9,999 steps daily
Active: More than 10,000 steps daily
Highly active: More than 12,500 steps daily
Resource:[link](https://www.medicinenet.com/how_many_steps_a_day_is_considered_active/article.htm)

High, or large, standard deviation indicates data are more spread out.

Now we can make calculation with sedentary_minutes measurement to find out how much time users spend passive.

```{r}
daily_activity$sedentary_minutes[daily_activity$sedentary_minutes == 0] <- NA

sedentary_info <- daily_activity %>% 
  summarize(min_sedentary = min(sedentary_minutes, na.rm = TRUE)/60,
            avg_sedentary = mean(sedentary_minutes, na.rm = TRUE)/60,
            medium_sedentary = median(sedentary_minutes, na.rm = TRUE)/60,
            max_sedentary = max(sedentary_minutes, na.rm = TRUE)/60)

sedentary_info
```
In this output, I see some problems, because it can't be true, that people in average sleep 7 hours, walk 8000 steps and sit 16 hours. I may conclude, that in this smart device sleep and sedentary time summarized. If we subtract from the average sedentary time sleep time : 16.5 - 6.99 = 9.51 , this seems more realistic, as in average we have 8 hours working day.

Let's count average step_total per month for every id in our list and include average sedentary time.

```{r}
# average steps by every id 
average_steps_per_month <- daily_steps %>%
  group_by(id) %>%
  summarise(avg_steps = mean(step_total))

# average sedentary time per month by every id 
average_sedentary_per_month <- daily_activity %>%
  group_by(id) %>%
  summarise(avg_sedentary = mean(sedentary_minutes))


#combine this two data sets to see is there any trends between this rates
average_steps_sedentary <- merge(average_sedentary_per_month, average_steps_per_month, by = "id", all.x = TRUE)

print(average_steps_sedentary)

```
Visualization of this data:

```{r}
ggplot(data = average_steps_sedentary) +
geom_point(mapping = aes(x = avg_steps, y = avg_sedentary/60))
```

**With this visualization we can find also some trends**
1. People who have high sedentary rate, tend to have less count of steps per day.
2. People who have lower sedentary time tend to have more activity in steps daily.


Now we upload another data set - Daily Intensities.
```{r}
intensities <- read_csv("dailyIntensities_merged.csv")
head(intensities)
```

Now we upload information about sleep - Sleep Day merged, to explore more trends analyzing this information.

```{r}

sleep_day <- read_csv("sleepDay_merged.csv")
sleep_day_df <- clean_names(sleep_day)
head(sleep_day_df)
```

Replacing all 0 with NA:
```{r}

sleep_day_df[sleep_day_df == 0] <- NA

```

Calculating average information about all users:

```{r}
sleep_info <- sleep_day_df %>% mutate(time_not_asleep = (total_time_in_bed - total_minutes_asleep)) %>% 
  summarize(min_sleep = min(total_minutes_asleep, na.rm = TRUE)/60,
            avg_sleep = mean(total_minutes_asleep, na.rm = TRUE)/60,
            medium_sleep = median(total_minutes_asleep, na.rm = TRUE)/60,
            max_sleep = max(total_minutes_asleep, na.rm = TRUE)/60,
            avg_time_not_asleep = mean(time_not_asleep))
sleep_info
```
In average users sleep 7 hours, and in average it takes 40 minutes in bed not sleeping. With given to us information we can't know , is those 40 minutes people lying in bed before being asleep, or after awakening. Let's assume, that this 40 minutes people can't fall asleep.
**Then we have another trend**:
People, that use smart device in average can't fall asleep for 40 minutes.


```{r}
sleep_day_df <- mutate(sleep_day_df, time_not_asleep = (total_time_in_bed - total_minutes_asleep))
sleep_day_df
```


```{r}
ggplot(data = sleep_day_df) +
  geom_point(mapping = aes(x = total_minutes_asleep, y = time_not_asleep )) + 
  geom_smooth(mapping = aes(x = total_minutes_asleep, y = time_not_asleep ))
```
In this visualization we can observe, that people who sleep between 6 to 10 hours daily tend to fall asleep faster (from 1 up to 50 minutes), then those , who sleep 3 - 5 hours.


Let's count users, who use another functions in smart device:



```{r}
weight_info <- read_csv("weightLogInfo_merged.csv")
calories_info <- read_csv("dailyCalories_merged.csv")
heartrate <- read_csv("heartrate_seconds_merged.csv")
```


```{r}
n_distinct(weight_info$Id)
n_distinct(daily_activity$id)
n_distinct(daily_steps$id)
n_distinct(intensities$Id)
n_distinct(sleep_day$Id)
n_distinct(calories_info$Id)
n_distinct(heartrate$Id)
```
Besides we have gotten trends from analyzing data sets closely, we can get information about using smart device from counting distinct id in data sets that shows us different activities.
so far, we can see those **trends**: 
1. Only 26.6% of smart device users used function that looks after weight.
2. Activity of all users was fixed by steps and calories indicator.
3. 24 of users was wearing smart device, while sleeping.
4. Smart device recorded information about heartbeat for 46% of users.

Let's look closely to information about hourly steps.
```{r}
hourly_steps1 <- read_csv("hourlySteps_merged.csv")
hourly_steps <- clean_names(hourly_steps1)
head(hourly_steps)
```


Now we can find average steps total for all users for the whole week 

```{r}
average_steps_by_day_sum_of_all <- tapply(hourly_steps$step_total, factor(weekdays(as.Date(hourly_steps$activity_hour))), FUN = mean)

average_steps_by_day <- (average_steps_by_day_sum_of_all)*24

days_of_week <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")

days_factor <- factor(names(average_steps_by_day), levels = days_of_week)

sorted_average_steps <- average_steps_by_day[order(days_factor)]

print(sorted_average_steps)


```

```{r}

days_of_week <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")

sorted_data <- data.frame(
  Day = factor(names(sorted_average_steps), levels = days_of_week), 
  Steps = sorted_average_steps
)


ggplot(sorted_data, aes(x = Day, y = Steps)) +
  geom_bar(stat = "identity", fill = "skyblue", width = 0.7) +
  labs(title = "Average Steps per Day of the Week", x = "Day", y = "Average Steps") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```
So, let's summarize our data in trends.
**Trends**:
1.	Only 26.6% of smart device users used function that looks after weight.
2.	Activity of all users was fixed by steps and calories indicator.
3.	24 of 33 users was wearing smart device, while sleeping
4.	Smart device recorded information about heartbeat for 46% of users.
5.	Some people count calories and doesn’t count step, and reverse.
**Trends among people, who use smart devices**:
1.	Some people count calories and doesn’t count step.
2.	People who have high sedentary rate, tend to have less count of steps per day.
3.	People who have lower sedentary time tend to have more activity in steps daily.
4.	People, that use smart device in average can't fall asleep for 40 minutes.
5.	People who sleep between 6 to 10 hours daily tend to fall asleep, then those, who sleep 3 - 5 hours.



